---
title: "Joining Data in R with dplyr"
subtitle: Lessons from DataCamp
output:
  pdf_document:
    toc: yes
    toc_depth: 3
  html_notebook:
    toc: yes
    toc_float: true
  html_document:
    toc: yes
    toc_float: true
---

#Introduction

The following document outlines the written portion of the lessons from [DataCamp](https://www.datacamp.com/)'s [Joining Data in R with dplyr](https://www.datacamp.com/courses/joining-data-in-r-with-dplyr). This requires Intermediate R-Knowledge and understanding of the dplyr package.

As a note: All text is completely copied and pasted from the course. There are instances where the document refers to the "editor on the right", please note, that in this notebook document all of the instances are noted in the "r-chunks" (areas containing working r-code), which occurs below the text, rather than to the right. Furthermore, This lesson contained instructional videos at the beginning of new concepts that are not detailed in this document. However, even without these videos, the instructions are quite clear in indicating what the code is accomplishing. 

<u>*If you have this document open on "R-Notebook", simply click "run" -> "Run all" (Or just press 'ctrl + alt + r'), let the "r-chunks" run (This might take a bit of time) then click "Preview". There are 5 necessary datasets to run this program, please create an r-project with this data or set a working directory (required files names are available in the "Required data for this session" section)*</u>

This document was created by Neil Yetz on 08/08/2018. Please send any questions or concerns in this document to Neil at <ndyetz@gmail.com>



##Required packages for this session

Below are the install.packages and libraries you will need to have in order to run this session successfully.

```{r, message = FALSE}
library(readr)
library(dplyr)

```

##Required data for this session

```{r, message = FALSE, warning = FALSE, results = 'hide'}

setwd("Data/")

#read all .csv files
temp = list.files(pattern="*.csv")
list2env(
  lapply(setNames(temp, make.names(gsub("*.csv$", "", temp))), 
         read_csv), envir = .GlobalEnv)
rm(temp)


#load R files
eighties <- readRDS("eighties.RDS")
lahmanNames <- readRDS("lahmanNames.RDS")
seventies <- readRDS("seventies.RDS")



```

#Course Description

This course builds on what you learned in Data Manipulation in R with dplyr by showing you how to combine data sets with dplyr's two table verbs. In the real world, data comes split across many data sets, but dplyr's core functions are designed to work with single tables of data. In this course, you'll learn the best ways to combine data sets into single tables. You'll learn how to augment columns from one data set with columns from another with mutating joins, how to filter one data set against another with filtering joins, and how to sift through data sets with set operations. Along the way, you'll discover the best practices for building data sets and troubleshooting joins with dplyr. Afterwards, you’ll be well on your way to data manipulation mastery!

#Chapter 1: Mutating joins

Mutating joins add new variables to one dataset from another dataset, matching observations across rows in the process. This chapter will explain the various ways you can join datasets together and what happens when you do.


##A basic join
As Garrett mentioned in the video, left_join() is the basic join function in dplyr. You can use it whenever you want to augment a data frame with information from another data frame.

For example, left_join(x, y) joins y to x. The second dataset you specify is joined to the first dataset. Keep that in mind as you go through the course.

For this exercise, the code in the editor sets up a left join but doesn't finish it. It's up to you to complete it.

INSTRUCTIONS

Complete the code to join artists to bands. bands2 should contain all of the information in bands supplemented with information in artists.

Print bands2 to the console to see the result.

```{r}
# Complete the code to join artists to bands
bands2 <- left_join(bands, artists, by = c("first", "last"))

# Examine the results
bands2
```


##A second join
You should always check the output of your joins. Just because a join worked, doesn't mean that it worked as you expected.

For example, the code in the editor joins the same two datasets that you joined in the previous exercise, but it returns a different result. Can you tell what went wrong?

INSTRUCTIONS

The result from the previous exercise, bands2, is loaded in your workspace.

Examine the output from the code provided in the editor. How is it different from bands2?
Fix the code so that the result is identical to bands2.

```{r}
# Fix the code to recreate bands2
left_join(bands, artists, by = c("first", "last"))
```

##A right join
There is more than one way to execute a left join. Knowing multiple methods will make you a more versatile data scientist, especially as you try to fit joins into pipes created with %>%.

In this exercise, you'll recreate bands2 once more, but this time without using a left_join().

INSTRUCTIONS

Use right_join() to create bands3, a new dataset that contains the same information as bands2.
Use setequal() to check that the datasets are the same.

```{r}
# Finish the code below to recreate bands3 with a right join
bands2 <- left_join(bands, artists, by = c("first", "last"))
bands3 <- right_join(artists, bands, by = c("first", "last"))

# Check that bands3 is equal to bands2
setequal(bands2, bands3)
```


##Inner joins and full joins
You may have noticed that some of the songs in songs correspond to some of the albums in albums. Suppose you want a new dataset that contains all of the songs for which you have data from both albums and songs. How would you make it?

The artists and bands datasets also share some information. What if you want to join these two datasets in such a way that you retain all of the information available in both tables, without throwing anything away?

You can think of inner joins as the most strict type of join: they only retain observations that appear in both datasets. In contrast, full joins are the most permissive type of join: they return all of the data that appears in both datasets (often resulting in many missing values).

Recall that, *_join(x, y) joins y to x. The second dataset you specify is joined to the first dataset.

INSTRUCTIONS

Join albums to songs in a way that returns only rows that contain information about both songs and albums.
Join bands to artists to create a single table that contains all of the available data.

```{r}
# Join albums to songs using inner_join()
inner_join(songs, albums, by = "album")

# Join bands to artists using full_join()
full_join(artists, bands, by = c("first", "last"))
```

##Pipes
You can combine dplyr functions together with the pipe operator, %>%, to build up an analysis step-by-step. %>% takes the result of the code that comes before it and "pipes" it into the function that comes after it as the first argument of the function.

So for example, the two pieces of code below do the same thing:

```
full_join(artists, bands, 
          by = c("first", "last"))

artists %>% 
  full_join(bands, by = c("first", "last"))
```
  
Pipes are so efficient for multi-step analysis that you will use them for the remainder of the exercises in this course. (If you need a refresher on the pipe operator, check out Data Manipulation in R with dplyr course!)

INSTRUCTIONS

The code in the editor finds all of the known guitarists in the bands dataset. Rewrite the code to use %>%s instead of multiple function calls. The pipe %>% should be used three times and temp zero times.


```{r}
# Find guitarists in bands dataset (don't change)
temp <- left_join(bands, artists, by = c("first", "last"))
temp <- filter(temp, instrument == "Guitar")
select(temp, first, last, band)

# Reproduce code above using pipes
bands %>% 
  left_join(artists, by = c("first", "last")) %>%
  filter(instrument == "Guitar") %>%
  select(first, last, band)
```


```{r, echo = FALSE}
#Ignore
goal <- artists %>%
  full_join(bands, by = c("first", "last")) %>%
  inner_join(songs, by = c("first", "last"))
```


##Practice with pipes and joins
We've created a data frame for you called goal. It's available in your workspace, so go ahead and take a look. Your mission, if you choose to accept it, is to create a dataset that's identical to goal.

INSTRUCTIONS

Examine the goal dataset by printing it to the console.
Write a pipe that uses a full join and an inner join to combine artists, bands, and songs into goal2, a dataset identical to goal.
Use setequal() to check that goal is identical to goal2.

```{r}
# Examine the contents of the goal dataset
goal

# Create goal2 using full_join() and inner_join() 
goal2 <- artists %>%
  full_join(bands, by = c("first", "last")) %>%
  inner_join(songs, by = c("first", "last"))
  
  
  
# Check that goal and goal2 are the same
setequal(goal, goal2)
```

##Choose your joins
You're getting the hang of pipes now! They are a very useful way to combine multiple joins to make a single dataset.

Let's craft one more dataset before moving to Chapter 2. One of the most useful ways to combine data is to place all of the values of all of the datasets into a single table.

INSTRUCTIONS

Write a pipe that combines artists, bands, songs, and albums (in that order) into a single table, such that it contains all of the information in the datasets.

```{r}
# Create one table that combines all information
artists %>%
  full_join(bands,  by = c("first", "last")) %>%
  full_join(songs,  by = c("first", "last")) %>%
  full_join(albums, by = c("album", "band"))
```



#Chapter 2: Filtering joins and set operations

Filtering joins and set operations combine information from datasets without adding new variables. Filtering joins filter the observations of one dataset based on whether or not they occur in a second dataset. Set operations use combinations of observations from both datasets to create a new dataset.

#Chapter 3: Assembling data

This chapter will show you how to build datasets from basic elements: vectors, lists, and individual datasets that do not require a join. dplyr contains a set of functions for assembling data that work more intuitively than base R's functions. The chapter will also look at when dplyr does and does not use data type coercion.

#Chapter 4: Advanced joining

Now that you have the basics, let's dive deep into the mechanics of joins. This chapter will show you how to spot common join problems, how to join based on multiple or mismatched keys, how to join multiple tables, and how to recreate dplyr's joins with SQL and base R.


#Chapter 5: Case study

You know the ins and outs of two-table verbs with dplyr, but your knowledge is untried! Let's cement what you've learned with a real world application.




